Local Type Inference for Polarised System F with Existentials

### 2023-11-21 oopsla24/tex/05-proof.tex

\subsection{Singularity}

@@

\paragraph{Mutual Induction of Soundness and Completeness}

The soundness and completeness of type inference cannot be proven separately:
the inductive proof of one requires another, and vice versa. 
Soundness proof requires completeness in the case of 
\ruleref{\ottdruleATAppLetLabel},
The soundness proof can be considered as a transformation from the algorithmic 
inference tree to a declarative one. In the case of \ruleref{\ottdruleATAppLetLabel},
the declarative rule requires that the (declaratively) inferred type of the application
is unique.  To show that, we take an arbitrary application typing tree and apply the typing
\emph{completeness}. Then the soundness of singularity

>>>>

\subsection{Singularity}
\label{sec:singularity-proof}

@@

\paragraph{Mutuality of the Soundness and Completeness Proofs}

Typically in our inductive proofs, the soundness is proven before completeness,
as the completeness requires certain properties of the premise subtrees.
However, in the case of the typing algorithm, the soundness and completeness proofs
cannot be separated: the inductive proof of one requires another, and vice versa.

The soundness proof requires completeness in the case of 
\ruleref{\ottdruleATAppLetLabel}, whose declarative counterpart requires \emph{uniqueness}
of the inferred application type. To show the uniqueness, we
need to conclude that any other \emph{declarative} tree 
must infer the same (or rather equivalent) type. 
The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proo}),
however, first, the declarative tree must be converted to the algorithmic one (by completeness!).

This way, these lemmas are proven by mutual induction. 
Since the declarative and the algorithmic systems are not mutually inductive,
we must introduce a metric on which the induction is conducted.

\paragraph{Inductive Metric}

Let us extend the declarative typing metric (\cref{def:decl-typing-metric}) 
to the algorithmic typing.

\begin{definition}[Number of prenex quantifiers]
    We define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of 
    prenex quantifiers, i.e.
    \begin{itemize}
        \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
        \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
    \end{itemize}
\end{definition}

\begin{definition}[Size of a Declarative Judgement]
    \label{def:decl-typing-size}
    For a declarative typing judgment $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
            (\size{[[args]]}, \npq{[[iN]]})$)
    \end{itemize}
    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
    syntax tree of the term $[[v]]$ or $[[c]]$
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
\end{definition}

\begin{definition}[Number of Equivalence Nodes]
    For a tree $T$ inferring
    a declarative typing judgment,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

\begin{definition}[Declarative Tree Metric]
    \label{def:decl-typing-metric}
    For a tree $T$ inferring
    a declarative typing judgment $J$,
    we define a metric $\metric{T}$
    as a pair $(\size{J}, \eqNodes{T})$.
\end{definition}

\begin{definition}[Size of an Algorithmic Judgement]
    For an algorithmic typing judgement $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊨ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊨ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]} = 
            (\size{[[args]]}, \npq{[[uN]]})$)
    \end{itemize}
\end{definition}

\begin{definition}[Algorithmic Tree Metric]
    For a tree $T$ inferring an algorithmic typing judgement $J$, we define 
    $\metric{T}$ as $(\size{J}, 0)$.
\end{definition}
